---
title: "Road Safety data Dashboard"
output: 
  flexdashboard::flex_dashboard:
    theme: flatly
    orientation: rows
    social: menu    
    source_code: embed
---

```{r setup, include=FALSE}
source("init.R")

accidents2018 <- read.csv("data/raw/accident2018") %>% 
  select(-X, -starts_with("location_"))

total_accidents2018 <- accidents2018 %>% 
  group_by(accident_severity) %>% 
  tally() 

total_acc18_fatal <- total_accidents2018$n[total_accidents2018$accident_severity == "Fatal"]
total_acc18_serious <- total_accidents2018$n[total_accidents2018$accident_severity == "Serious"]
total_acc18_slight <- total_accidents2018$n[total_accidents2018$accident_severity == "Slight"]
total_acc18 <- total_acc18_fatal + total_acc18_serious + total_acc18_slight

# Variables.
color_total <- "#1F77B4"
color_serious <- "#FF7F0E"
color_uci <- "pink"
color_slight <- "#2C9F2C"
color_fatal <- "rgba(214, 39, 40, 1)"

```

menu item 1
=====================================  
Row
------------------
### Total {.value-box .no-mobile}

```{r}

valueBox(value = format(total_acc18, big.mark = ","),
         caption = paste0("Total recorded accidents in UK during 2018"), 
   icon = "fa fa-car-crash", 
   color = color_total)
```

### Fatal {.value-box .no-mobile}

```{r}

valueBox(value = format(total_acc18_fatal, big.mark = ","),
         caption = paste0("Total fatal accidents, which account to <b>",
                          round(total_acc18_fatal/total_acc18*100,  
                          digits = 2),
                    "% of total accidents.</b> <small>Year: 2018</small>"), 
   icon = "fab fa-skull-crossbones", 
   color = color_fatal)
```

### Serious{.value-box .no-mobile

```{r}

valueBox(value = format(total_acc18_serious, big.mark = ","),
         caption = paste0("Total serious accidents, which account to <b>",
                          round(total_acc18_serious/total_acc18*100,  
                          digits = 2),
                    "% of total accidents.</b> <small>Year: 2018</small>"), 
   icon = "fas fa-ambulance", 
   color = color_serious)
```

### Slight {.value-box .no-mobile}

```{r}

valueBox(value = format(total_acc18_slight, big.mark = ","),
         caption = paste0("Total slight accidents, which account to <b>",
                          round(total_acc18_slight/total_acc18*100,  
                          digits = 2),
                    "% of total accidents.</b> <small>Year: 2018</small>"), 
   icon = "fas fa-user-injured", 
   color = color_slight)
```


Row {.tabset}
-----------------------------------------------------------------------

### Accidents by location and type 

```{r}
summary_severity = accidents2018 %>% 
  select(accident_severity) %>%
  group_by(accident_severity) %>%
  summarise(Total = n()) %>%
  mutate(freq = Total/sum(Total)*100) %>%
  arrange(desc(Total))

ggplot(summary_severity, aes(x = reorder(accident_severity, Total), 
                                     y = Total)) +
  geom_bar(stat = "identity") +
  ggtitle("Total cases by accident severity") +
  labs(x = "Accident severity", y = "Total cases") +
  coord_flip() +
  geom_text(aes(label = Total), size = 3, hjust = -.5, vjust = .5, position = "stack") +
  scale_fill_brewer(type = "seq", palette = "Reds") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_blank()) +
  guides(fill = guide_legend(title.position = "left", nrow=1, byrow=TRUE))
```

```{r}
ggplot(data = accidents2018, aes(x = urban_or_rural_area, fill = accident_severity)) +
  geom_bar()

ggplotly()
```

URban and rural areas follow the same patterns in severity type. 


```{r}

spdf <- accidents2018 %>%
  select(latitude, longitude, accident_severity) %>% 
  mutate(accident_severity = as.factor(accident_severity)) %>% 
  filter(!is.na(latitude))

leaflet(options = leafletOptions(minZoom = 0, maxZoom = 14)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addHeatmap(data = filter(spdf, accident_severity == "Fatal"), group = "Fatal Accidents",
             lng = ~longitude, lat = ~latitude,
             blur = 20, max = 0.4, radius = 10 ) %>%
  addHeatmap(data = filter(spdf, accident_severity == "Serious"), group = "Serious Accidents",
           lng = ~longitude, lat = ~latitude,
           blur = 20, max = 0.4, radius = 10 ) %>%
  addHeatmap(data = filter(spdf, accident_severity == "Slight"), group = "Slight Accidents",
           lng = ~longitude, lat = ~latitude,
           blur = 20, max = 0.4, radius = 10 ) %>%
  addMiniMap(tiles = providers$CartoDB.Positron) %>% 
  addLayersControl(
    # baseGroups = c("OSM (default)", "Toner", "Toner Lite"),
    overlayGroups = c("Fatal Accidents", "Serious Accidents", "Slight Accidents"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% 
  hideGroup(c("Serious Accidents", "Slight Accidents")) %>% 
  addControl("Accidents by location and type", position = "topleft", className="map-title")

```



### Explore the data

```{r}
datatable(head(accidents2018))
```


About
=====================================  

#### About this dashboard

This dashboard is a proof of concept and fast-prototype to explore STATS-19 data and just for testing purposes and exploring posibilities. 

* Source code: https://github.com/ccamara/road_safety_data/
* Author: [Carlos Cámara](https://carloscamara.es)
* Licence: MIT

It relies in the following packages:

Iannone, R., Allaire, J., & Borges, B. (2018). flexdashboard: R markdown format for flexible dashboards [Manual]. https://CRAN.R-project.org/package=flexdashboard

Lovelace, R., Morgan, M., Hama, L., Padgham, M., Ranzolin, D., & Sparks, A. (2019). stats 19: A package for working with open road crash data. The Journal of Open Source Software, 4(33), 1181. https://doi.org/10.21105/joss.01181

Sievert, C. (2020). Interactive web-based data visualization with r, plotly, and shiny. Chapman and Hall/CRC. https://plotly-r.com

Wickham, H., François, R., Henry, L., & Müller, K. (2020). dplyr: A grammar of data manipulation [Manual]. https://CRAN.R-project.org/package=dplyr
